Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2014-08-26 paper.tex
Added quite a bit of DD theory to the introduction section.

The first three-dimensional simulations of double degenerate systems were performed by \citet{benz:1990}, who used the smoothed particle hydrodynamics (SPH) method to simulate the merger process. They found that if the lower-mass secondary was close enough to the primary to begin mass transfer on a dynamical time scale, the secondary completely disrupted and formed a hot envelope on the primary, with an accretion disk surrounding the core and envelope. They noted that the envelope temperature was hot enough to generate carbon fusion but neglected reactions in their simulation. Later simulations of similar setups \citep{rasio_shapiro:1995,yoon:2007} corroborated this result. This validation of the prediction of \cite{tutukov_yungelson:1979} did not instill confidence in the community. \cite{mochkovitch_livio:1990} and \cite{livio:2000} observed that turbulent viscosity would be sufficiently large for angular momentum to be removed from the disk at a rate high enough to generate these troublesome accretion timescales. Based on all of this evidence, the review of \cite{hillebrandtniemeyer2000} argued that the model was only viable if the accretion-induced collapse problem could be avoided. Later work looked more carefully at the initial conditions for the simulation and reached the conclusion that mass transfer from the secondary would occur over many orbits, instead of disrupting completely over the span of a few orbits as originally suggested (ref). This may be a way to avoid the off-center carbon ignition (ref).

The most recent shift in perspective on this problem started with \cite{pakmor:2010}. This group used the SPH method to study the merger of equal-mass ($0.9\ M_\odot$) white dwarfs and found that a hotspot was generated near the surface of the primary white dwarf. They argued that this region had a temperature and density sufficient to trigger a thermonuclear detonation. They propagated this detonation throughout the system and found that it would observationally appear as a subluminous Type Ia supernova. This was the first time any simulation successfully reproduced at least some characteristics of a SN Ia. \cite{pakmor:2011} tried a few different mass combinations and found empirically that this would hold as long as the secondary was at least 80\% as massive as the primary. \cite{pakmor:2012} followed this with a study of a $1.1\ M_\odot$ and $0.9\ M_\odot$ pair, with more careful attention to the initial conditions of the simulation, and found that they could reproduce the observational characteristics of normal SNe Ia. \cite{pakmor:2013} refined this approach by adding a thin helium shell on the primary, and found that this robustly led to a detonation of the white dwarf. They argued that this framework could explain most of the observed types of SNe Ia because it allowed for both carbon-oxygen and helium white dwarfs to be the secondary, allowing for a wide range of possible outcomes.

% Discuss some other recent work (Guillochon et al. 2010, van Kerkwijk et al. 2012)

>>>>

The first three-dimensional simulations of double degenerate systems were performed by \citet{benz:1990}, who used the smoothed particle hydrodynamics (SPH) method to simulate the merger process. They found that if the lower-mass secondary was close enough to the primary to begin mass transfer on a dynamical time scale, the secondary completely disrupted and formed a hot envelope on the primary, with a centrifugally-supported accretion disk surrounding the core and envelope. They noted that the envelope temperature was hot enough to generate carbon fusion but neglected reactions in their simulation. Later simulations of similar setups \citep{rasio_shapiro:1995,yoon:2007,loren-aguilar:2009} corroborated this result. This validation of the prediction of \cite{tutukov_yungelson:1979} did not instill confidence in the community. \cite{mochkovitch_livio:1990} and \cite{livio:2000} observed that turbulent viscosity would be sufficiently large for angular momentum to be removed from the disk at a rate high enough to generate these troublesome accretion timescales. Based on all of this evidence, the review of \cite{hillebrandtniemeyer2000} argued that the model was only viable if the accretion-induced collapse problem could be avoided. Later work by \cite{shen:2012} and \cite{schwab:2012} used a more detailed treatment of the viscous transport in the outer regions of the remnant and found that while the centrifugally supported envelope would be converted into hot envelope material on a viscous timescale, their simulations still led to off-center carbon burning. \cite{vankerkwijk:2010} argued that equal-mass mergers would lead to the conditions necessary for carbon detonation in the center of the merged object, but \cite{shen:2012} questioned this for similar reasons related to how viscous transport would convert rotational motion into pressure support. \cite{zhu:2013} followed this with an expanded parameter space study and argued that many of their carbon-oxygen systems had the potential to detonate. The study of the long-term evolution of the remnants is thus still an open subject of research.

The most recent shift in perspective on this problem started with two series of papers appearing at roughly the same time. One began with \cite{pakmor:2010} and \cite{pakmor:2011}. This group used the SPH method to study the merger of equal-mass ($0.9\ M_\odot$) carbon-oxygen white dwarfs and found that a hotspot was generated near the surface of the primary white dwarf. They argued that this region had a temperature and density sufficient to trigger a thermonuclear detonation. They propagated this detonation throughout the system and found that it would observationally appear as a subluminous Type Ia supernova. This was the first time any simulation successfully reproduced at least some characteristics of a SN Ia. \cite{pakmor:2011} tried a few different mass combinations and found empirically that this would hold as long as the secondary was at least 80\% as massive as the primary. These events, where the merger process resulted in the detonation of the system during the merger coalescence -- avoiding the much longer time-scale evolution -- were termed ``violent'' mergers.

Around the same time, however, \cite{guillochon:2010} and \cite{dan:2011} came to significantly different conclusions. They pointed out that the previously mentioned simulations generally all shared a significant drawback, which was that their initial conditions were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006}, and \cite{motl:2007} (the first three-dimensional grid-based simulations of white dwarf mergers) pioneered the study of looking at the long-term dynamical evolution of merger scenarios after constructing exact equilibrium initial conditions. In contrast to the other studies, they found that paying close attention to the accuracy of the initial conditions of the simulation has important effects. In particular, earlier work placed the stars too close together and ignored the effects of tidal forces on changing the shape of the secondary, leading to the merger happening artificially too quickly. When the initial conditions are constructed in exact equilibrium, the system can be stable for tens of orbital periods, substantially changing the character of the mass transfer phase. However, one limitation of this work is that the authors used a polytropic equation of state and thus could not consider nuclear reactions. \cite{guillochon:2010} and \cite{dan:2011} improved on this using a realistic equation of state, a nuclear reaction network, and a similar approach to the equilibrium initial conditions, and found substantial agreement with the idea that mass transfer occurs in a stable manner over tens of orbital periods. They also found that, assuming the material accreted onto the surface of the primary was primarily helium, explosive surface detonations would occur as a result of accretion stream instabilities during the mass transfer phase prior to the full merger. This could trigger a double-detonation explosion and thus perhaps a SN Ia.

The most recent developments give some areas of agreement and some of remaining tension. \cite{pakmor:2012} performed a merger scenario with better treatment of the initial conditions and also found that the merger process happened over more than ten orbits. Nevertheless, they still found that a carbon-oxygen detonation would occur, in line with their earlier results. On the other hand, \cite{dan:2012} and \cite{dan:2014} performed a large sweep of the parameter space for merger pairs and found that pure carbon-oxygen systems would generally not lead to detonations and violent mergers. They did find that for systems containing helium, many of them would detonate and potentially lead to SNe Ia, either through the aforementioned instabilities in the accretion stream, or during the contact phase, similar to the violent carbon-oxygen mergers. \cite{pakmor:2013} added a thin helium shell on their primary white dwarf, and found that this robustly led to a detonation of the white dwarf. Thus the preliminary consensus seems to be that systems containing helium could robustly lead to events resembling SNe Ia, while there is still debate about larger-mass systems containing only carbon and oxygen.

% Still need Raskin/Moll/et al. papers
